---
title: "Week 15: Introduction to Time Series Analysis"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
---

# Today's plan

1. Brief framing: what makes time series data special
2. Getting environmental data with `dataRetrieval`
3. The `tsibble`: a tidy time series object
4. Visualization (several kinds)
5. Decomposition: trend, seasonality, remainder
6. A simple model and forecast
7. Your turn

---

# Packages

Today we're using the **tidyverts** ecosystem — a tidyverse-native set of packages for time series work developed by Rob Hyndman's group.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(dataRetrieval)  # USGS water data
library(tsibble)        # tidy time series data structure
library(feasts)         # visualization + decomposition
library(fable)          # forecasting models
library(lubridate)
```

---

# What makes time series data special?

Most statistical methods assume observations are **independent** of each other. Time series data violates this: yesterday's river flow strongly predicts today's river flow. This dependence is called **autocorrelation**, and it's both a challenge and a tool.

We'll try to decompose any time series into three components:

- **Trend** — long-term increase or decrease
- **Seasonality** — regular, repeating patterns (e.g., annual cycles)
- **Remainder** — what's left after accounting for trend and seasonality

Think of it like: *data = trend + seasonality + remainder*

### Quick discussion

Before we look at any data: if you were measuring streamflow on a river in Northeast Ohio, what kind of trend and seasonal pattern would you expect? Why?

---

# Getting data: USGS streamflow

The `dataRetrieval` package provides direct access to USGS water data. We'll look at streamflow on the **Cuyahoga River at Independence, OH** — just downstream from us.

```{r}
site_no      <- "04208000"   # Cuyahoga River at Independence, OH
parameter_cd <- "00060"      # discharge, in cubic feet per second (cfs)

raw_flow <- dataRetrieval::readNWISdv(
  siteNumbers = site_no,
  parameterCd = parameter_cd,
  startDate   = "2000-01-01",
  endDate     = "2023-12-31"
)

head(raw_flow)
```

**Question**: What variables do you see? What do you think `X_00060_00003` means? What about the `_cd` column?

```{r}
# The package has a helper to rename columns to something readable
flow_clean <- raw_flow %>%
  dataRetrieval::renameNWISColumns() %>%
  dplyr::select(Date, Flow) %>%
  dplyr::filter(!is.na(Flow))

head(flow_clean)
```

How many observations do we have?

```{r}
nrow(flow_clean)
```

**Question**: Is that about what you'd expect for daily data from 2000–2023? Do the math.

---

# The `tsibble`: a tidy time series object

A **tsibble** is like a tibble, but with an explicit time **index** and (optionally) a grouping **key**. The index tells the package how observations are ordered in time and infers the interval automatically.

```{r}
flow_ts <- flow_clean %>%
  tsibble::as_tsibble(index = Date)

flow_ts
```

**Question**: What does tsibble report for the interval? Does that match what we pulled?

---

# Visualization

## The basic time series plot

```{r}
flow_ts %>%
  feasts::autoplot(Flow) +
  labs(
    title = "Daily Streamflow — Cuyahoga River at Independence, OH",
    y     = "Discharge (cfs)",
    x     = NULL
  )
```

What do you notice? Any obvious trend? Seasonality? Large spikes?

## Aggregate to monthly

Daily data for 24 years is noisy. Monthly means will make seasonal patterns clearer and keep our models fast.

```{r}
flow_monthly <- flow_clean %>%
  dplyr::mutate(ym = tsibble::yearmonth(Date)) %>%
  dplyr::group_by(ym) %>%
  dplyr::summarise(mean_flow = mean(Flow, na.rm = TRUE)) %>%
  tsibble::as_tsibble(index = ym)

flow_monthly
```

```{r}
flow_monthly %>%
  feasts::autoplot(mean_flow) +
  labs(
    title = "Monthly Mean Streamflow — Cuyahoga River at Independence, OH",
    y     = "Mean Discharge (cfs)",
    x     = NULL
  )
```

How does this compare to the daily plot? What's easier or harder to see at monthly resolution?

## Seasonal plot

This plots each year as its own line, all aligned on the same x-axis (month of year).

```{r}
flow_monthly %>%
  feasts::gg_season(mean_flow, labels = "both") +
  labs(
    title = "Seasonal Plot — Monthly Mean Streamflow",
    y     = "Mean Discharge (cfs)"
  )
```

**Questions**:

1. Which months consistently have the highest flow? Lowest?
2. How consistent is the seasonal pattern across years?
3. Can you identify any years that look unusual? What might explain them?

## Subseries plot

Another way to look at seasonality: one panel per month, showing how that month's values change across years.

```{r}
flow_monthly %>%
  feasts::gg_subseries(mean_flow) +
  labs(
    title = "Subseries Plot — Monthly Mean Streamflow",
    y     = "Mean Discharge (cfs)"
  )
```

The blue horizontal line in each panel is the mean for that month. What does this view tell you that the seasonal plot didn't?

## Autocorrelation

The ACF (Autocorrelation Function) measures how correlated each observation is with its own past values — at different lag distances.

```{r}
flow_monthly %>%
  feasts::ACF(mean_flow, lag_max = 36) %>%
  feasts::autoplot() +
  labs(title = "Autocorrelation Function — Monthly Mean Streamflow")
```

**Questions**:

1. What do the blue dashed lines represent?
2. At lag 1 the correlation is strongly positive — what does that mean in plain language?
3. What pattern do you see at lags 12, 24, 36? Why does that make sense given the data?

---

# Decomposition

STL (Seasonal and Trend decomposition using Loess) separates our time series into its component parts. It's robust to outliers and flexible about the shape of the trend and seasonal components.

```{r}
dcmp <- flow_monthly %>%
  fabletools::model(
    stl = feasts::STL(mean_flow ~ trend(window = 21) + season(window = "periodic"))
  )

dcmp %>%
  fabletools::components() %>%
  feasts::autoplot() +
  labs(title = "STL Decomposition — Monthly Mean Streamflow")
```

Look at each panel:

- **data** — original series
- **trend** — long-term direction (smoothed out the seasonality)
- **season_year** — repeating annual cycle
- **remainder** — what's left over after removing trend and season

**Questions**:

1. Is there a long-term trend in streamflow over this period? What direction?
2. How large is the seasonal component relative to the trend? (Hint: look at the y-axis scales)
3. The grey bars on the right side of each panel are deliberately the same height — what is their purpose?
4. What might explain large spikes in the remainder component?

---

# A simple model and forecast

We'll use a **Time Series Linear Model** (`TSLM`). Think of it as `lm()` but aware of the time structure. We'll include a linear trend term and monthly seasonal dummies.

```{r}
fit <- flow_monthly %>%
  fabletools::model(
    tslm = fable::TSLM(mean_flow ~ trend() + season())
  )

tidy(fit)
```

**Questions**:

1. What does the `trend()` coefficient tell you? What are its units?
2. There are seasonal coefficients — how would you interpret the one for `season()year2` (February)?
3. The model has a reference category for season — which month is it, and how can you tell?

## Model fit: observed vs. fitted

```{r}
fit %>%
  fabletools::augment() %>%
  ggplot(aes(x = ym)) +
  geom_line(aes(y = mean_flow, color = "Observed")) +
  geom_line(aes(y = .fitted,   color = "Fitted")) +
  scale_color_manual(values = c("Observed" = "gray50", "Fitted" = "steelblue")) +
  labs(
    title = "Observed vs. Fitted — TSLM",
    y     = "Mean Discharge (cfs)",
    x     = NULL,
    color = NULL
  )
```

Where does the model fit well? Where does it struggle? What does that tell you about the model's assumptions?

## Forecast

```{r}
fc <- fit %>%
  fabletools::forecast(h = "2 years")

fc %>%
  fabletools::autoplot(flow_monthly, level = c(80, 95)) +
  labs(
    title = "2-Year Forecast — Monthly Mean Streamflow",
    y     = "Mean Discharge (cfs)",
    x     = NULL
  )
```

The shaded bands are **prediction intervals** (80% and 95%).

**Questions**:

1. Does the forecast look reasonable? Does it match your intuition from the seasonal plot?
2. What does the *width* of the prediction intervals tell you?
3. What important things does this model NOT account for? What are its real-world limitations?

---

# Your turn

Choose one of the following tasks with the remaining time. Compare your results with classmates.

---

**Option 1 — Try a different gauge**

Find a USGS site number for a river you're interested in (use the [USGS Water Resources site](https://waterdata.usgs.gov/nwis) to find a site). Pull the data, convert to a monthly tsibble, and recreate at minimum: the seasonal plot, ACF, and decomposition.

Some Ohio gauges to try if you don't want to search:

| Site | Description |
|------|-------------|
| `04208504` | Cuyahoga River at Kent, OH |
| `04212100` | Grand River at Painesville, OH |
| `04208600` | Mahoning River at Youngstown, OH |

---

**Option 2 — Try a different model**

Go back to the monthly Cuyahoga data. Fit an ARIMA model instead of TSLM:

```{r eval=FALSE}
fit_arima <- flow_monthly %>%
  fabletools::model(
    arima = fable::ARIMA(mean_flow)
  )
```

Produce the same observed vs. fitted and forecast plots. How do the two models compare? Which do you prefer and why?

---

**Option 3 — Dig into the trend**

The decomposition suggested a possible trend. Is it statistically meaningful? Look at the `tidy(fit)` output again. What is the trend coefficient, its standard error, and its p-value? Write 2-3 sentences interpreting the result for a non-technical audience.

---

# Wrapping up

Today we covered:

- **`tsibble`** — the tidy time series data structure
- **Visualization** — time plots, seasonal plots, subseries plots, ACF
- **Decomposition** — separating trend, seasonality, and remainder with STL
- **Modeling** — fitting a simple trend + seasonality model and forecasting

The tidyverts ecosystem (`tsibble` + `feasts` + `fable`) has a lot more to offer. If you want to go deeper, Rob Hyndman's free textbook [Forecasting: Principles and Practice (3rd ed.)](https://otexts.com/fpp3/) is the best resource and uses exactly these packages.
