---
title: "Week 15: Introduction to Time Series Analysis"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
---

# Today's plan

1. Brief framing: what makes time series data special
2. Getting environmental data with `dataRetrieval`
3. The `tsibble`: a tidy time series object
4. Visualization (several kinds)
5. Decomposition: trend, seasonality, remainder
6. A simple model and forecast
7. Your turn

---

# Packages

Today we're using the **tidyverts** ecosystem — a tidyverse-native set of packages for time series work developed by Rob Hyndman's group.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(dataRetrieval)  # USGS water data
library(tsibble)        # tidy time series data structure
library(feasts)         # visualization + decomposition
library(fable)          # forecasting models
library(lubridate)
```

---

# What makes time series data special?

Most statistical methods assume observations are **independent** of each other. Time series data violates this: yesterday's river flow strongly predicts today's river flow. This dependence is called **autocorrelation**, and it's both a challenge and a tool.

We'll try to decompose any time series into three components:

- **Trend** — long-term increase or decrease
- **Seasonality** — regular, repeating patterns (e.g., annual cycles)
- **Remainder** — what's left after accounting for trend and seasonality

Think of it like: *data = trend + seasonality + remainder*

### Quick discussion

Before we look at any data: if you were measuring streamflow on a river in Northeast Ohio, what kind of trend and seasonal pattern would you expect? Why?

> **Likely answers:** For seasonality, expect high flows in late winter and spring (snowmelt + spring rainfall in March–May) and low flows in late summer (August–September, when evapotranspiration is high and precipitation is more sporadic). For trend, it's genuinely hard to predict without looking — flows might be increasing due to more intense precipitation events, decreasing due to water withdrawals or land-use change, or roughly flat. This is a good reminder that intuition should always be checked against data.

---

# Getting data: USGS streamflow

The `dataRetrieval` package provides direct access to USGS water data. We'll look at streamflow on the **Cuyahoga River at Independence, OH** — just downstream from us.

```{r}
site_no      <- "04208000"   # Cuyahoga River at Independence, OH
parameter_cd <- "00060"      # discharge, in cubic feet per second (cfs)

raw_flow <- dataRetrieval::readNWISdv(
  siteNumbers = site_no,
  parameterCd = parameter_cd,
  startDate   = "2000-01-01",
  endDate     = "2023-12-31"
)


head(raw_flow)
```

**Question**: What variables do you see? What do you think `X_00060_00003` means? What about the `_cd` column?

The column name encodes USGS codes: `00060` = discharge parameter, `00003` = daily mean statistic. The `_cd` column is a data-quality flag. Let's look at what values it actually contains:

```{r}
raw_flow %>% count(X_00060_00003_cd)
```

`"A"` = approved (reviewed), `"P"` = provisional (subject to revision). For long-term analysis, approved data is preferable.

```{r}
# The package has a helper to rename columns to something readable
flow_clean <- raw_flow %>%
  dataRetrieval::renameNWISColumns() %>%
  dplyr::select(Date, Flow) %>%
  dplyr::filter(!is.na(Flow))

head(flow_clean)
```

How many observations do we have?

```{r}
nrow(flow_clean)
```

**Question**: Is that about what you'd expect for daily data from 2000–2023? Do the math.

```{r}
# expected days in our date range
expected_days <- as.numeric(as.Date("2023-12-31") - as.Date("2000-01-01")) + 1
expected_days

# how does that compare to what we actually have?
nrow(flow_clean)

# how many NAs were dropped during cleaning?
raw_flow %>% dataRetrieval::renameNWISColumns() %>% pull(Flow) %>% is.na() %>% sum()
```

Any shortfall between expected and actual is missing data — equipment outages, ice affecting the sensor, etc.

---

# The `tsibble`: a tidy time series object

A **tsibble** is like a tibble, but with an explicit time **index** and (optionally) a grouping **key**. The index tells the package how observations are ordered in time and infers the interval automatically.

```{r}
flow_ts <- flow_clean %>%
  tsibble::as_tsibble(index = Date)

flow_ts
```

The header line of the output above shows the interval tsibble detected — look for `[1D]`. Yes, that matches: we pulled daily values. Tsibble infers the interval automatically from the spacing of the index. If you had gaps, it would error out rather than silently skip them — which is actually useful.

---

# Visualization

## The basic time series plot

```{r}
flow_ts %>%
  feasts::autoplot(Flow) +
  labs(
    title = "Daily Streamflow — Cuyahoga River at Independence, OH",
    y     = "Discharge (cfs)",
    x     = NULL
  )
```

What do you notice? Any obvious trend? Seasonality? Large spikes?

The large upward spikes are flood events. There's a seasonal rhythm, but it's hard to characterize precisely at daily resolution — the noise drowns it out. A long-term trend, if present, is essentially invisible here. We'll aggregate to monthly to make the signal easier to see.

## Aggregate to monthly

Daily data for 24 years is noisy. Monthly means will make seasonal patterns clearer and keep our models fast.

```{r}
flow_monthly <- flow_clean %>%
  dplyr::mutate(ym = tsibble::yearmonth(Date)) %>%
  dplyr::group_by(ym) %>%
  dplyr::summarise(mean_flow = mean(Flow, na.rm = TRUE)) %>%
  tsibble::as_tsibble(index = ym)

flow_monthly
```

```{r}
flow_monthly %>%
  feasts::autoplot(mean_flow) +
  labs(
    title = "Monthly Mean Streamflow — Cuyahoga River at Independence, OH",
    y     = "Mean Discharge (cfs)",
    x     = NULL
  )
```

How does this compare to the daily plot? What's easier or harder to see at monthly resolution?

The seasonal pattern is much clearer — regular annual peaks and troughs are no longer buried in noise. The trade-off: individual flood events are averaged away. For understanding long-term structure (seasonality, trend), monthly is appropriate. For flood frequency or magnitude, you'd stay with daily.

## Seasonal plot

This plots each year as its own line, all aligned on the same x-axis (month of year).

```{r}
flow_monthly %>%
  feasts::gg_season(mean_flow, labels = "both") +
  labs(
    title = "Seasonal Plot — Monthly Mean Streamflow",
    y     = "Mean Discharge (cfs)"
  )
```

**Questions**:

1. Which months consistently have the highest flow? Lowest? Can you tell?
2. How consistent is the seasonal pattern across years?
3. Can you identify any years that look unusual? What might explain them?

For question 1, let's compute this directly instead of just eyeballing:

```{r}
flow_monthly %>%
  as_tibble() %>%
  mutate(month = lubridate::month(ym, label = TRUE)) %>%
  group_by(month) %>%
  summarise(avg_flow = mean(mean_flow, na.rm = TRUE)) %>%
  arrange(desc(avg_flow))
```

For question 3, which individual years had the highest and lowest *annual* mean flow?

```{r}
flow_monthly %>%
  as_tibble() %>%
  mutate(year = lubridate::year(ym)) %>%
  group_by(year) %>%
  summarise(annual_mean = mean(mean_flow, na.rm = TRUE)) %>%
  arrange(desc(annual_mean))
```

Years at the top and bottom of this table are the ones that stand out as lines in the seasonal plot. The seasonal shape stays consistent year to year — the peak is almost always spring and the trough late summer — but the overall magnitude varies substantially.

## Subseries plot

Another way to look at seasonality: one panel per month, showing how that month's values change across years.

```{r}
flow_monthly %>%
  feasts::gg_subseries(mean_flow) +
  labs(
    title = "Subseries Plot — Monthly Mean Streamflow",
    y     = "Mean Discharge (cfs)"
  )
```

The blue horizontal line in each panel is the mean for that month. What does this view tell you that the seasonal plot didn't?

The subseries plot lets you ask: within a specific month, is flow changing over time? The seasonal plot shows the annual shape well but makes month-specific trends invisible. Here's one way to make that question explicit:

```{r}
# Is March flow trending over time?
flow_monthly %>%
  as_tibble() %>%
  mutate(month = lubridate::month(ym),
         year  = lubridate::year(ym)) %>%
  filter(month == 3) %>%
  ggplot(aes(x = year, y = mean_flow)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "March Mean Streamflow Over Time",
       y = "Mean Discharge (cfs)", x = "Year")
```

Try changing `month == 3` to another month. Does every month show the same direction of change?

## Autocorrelation

The ACF (Autocorrelation Function) measures how correlated each observation is with its own past values — at different lag distances.

```{r}
flow_monthly %>%
  feasts::ACF(mean_flow, lag_max = 36) %>%
  feasts::autoplot() +
  labs(title = "Autocorrelation Function — Monthly Mean Streamflow")
```

**Questions**:

1. What do the blue dashed lines represent?
2. At lag 1 the correlation is strongly positive — what does that mean in plain language?
3. What pattern do you see at lags 12, 24, 36? Why does that make sense given the data?

For question 1, the threshold is approximately ±2/√n. Let's compute it:

```{r}
n_months <- nrow(flow_monthly)
significance_threshold <- 2 / sqrt(n_months)
significance_threshold
```

Bars extending beyond this value are statistically distinguishable from zero — not just noise.

For question 3, we can extract the ACF values directly and look at the seasonal lags:

```{r}
flow_monthly %>%
  feasts::ACF(mean_flow, lag_max = 36) %>%
  as_tibble() %>%
  filter(as.numeric(lag) %in% c(1, 12, 24, 36))
```

The strong positive values at lags 12, 24, and 36 confirm annual seasonality: this month's flow is similar to the same month one, two, and three years ago. At lag 1, the positive correlation means high-flow months tend to follow high-flow months — the river has inertia, it doesn't jump randomly month to month.

---

# Decomposition

STL (Seasonal and Trend decomposition using Loess) separates our time series into its component parts. It's robust to outliers and flexible about the shape of the trend and seasonal components.

```{r}
dcmp <- flow_monthly %>%
  fabletools::model(
    stl = feasts::STL(mean_flow ~ trend(window = 21) + season(window = "periodic"))
  )

dcmp %>%
  fabletools::components() %>%
  feasts::autoplot() +
  labs(title = "STL Decomposition — Monthly Mean Streamflow")
```

Look at each panel:

- **data** — original series
- **trend** — long-term direction (smoothed out the seasonality)
- **season_year** — repeating annual cycle
- **remainder** — what's left over after removing trend and season

**Questions**:

1. Is there a long-term trend in streamflow over this period? What direction?
2. How large is the seasonal component relative to the trend? (Hint: look at the y-axis scales)
3. The grey bars on the right side of each panel are deliberately the same height — what is their purpose?
4. What might explain large spikes in the remainder component?

For questions 1 and 2, we can extract the decomposition components and compare their ranges directly:

```{r}
dcmp_components <- dcmp %>% fabletools::components()

dcmp_components %>%
  as_tibble() %>%
  summarise(
    trend_range    = diff(range(trend,       na.rm = TRUE)),
    seasonal_range = diff(range(season_year, na.rm = TRUE)),
    remainder_range= diff(range(remainder,   na.rm = TRUE))
  )
```

Whichever range is largest dominates the variation in the series. In most river data, the seasonal range dwarfs the trend — knowing what month it is tells you more about flow than knowing what year it is.

For question 4, which dates have the largest remainder values (biggest departures from trend + season)?

```{r}
dcmp_components %>%
  as_tibble() %>%
  select(ym, remainder) %>%
  arrange(desc(abs(remainder))) %>%
  head(10)
```

Look up those dates. They likely correspond to major flood events, drought years, or other unusual conditions. The grey bars in the decomposition plot serve as a visual scale reference: since all bars are the same physical height but span different y-axis ranges, you can visually compare how much variation each component actually accounts for.

---

# A simple model and forecast

We'll use a **Time Series Linear Model** (`TSLM`). Think of it as `lm()` but aware of the time structure. We'll include a linear trend term and monthly seasonal dummies.

```{r}
fit <- flow_monthly %>%
  fabletools::model(
    tslm = fable::TSLM(mean_flow ~ trend() + season())
  )

tidy(fit)
```

**Questions**:

1. What does the `trend()` coefficient tell you? What are its units?
2. There are seasonal coefficients — how would you interpret the one for `season()year2` (February)?
3. The model has a reference category for season — which month is it, and how can you tell?

```{r}
# Pull the trend coefficient and express it as monthly and annual change
trend_coef <- tidy(fit) %>% filter(term == "trend()")
trend_coef

# annual rate of change
trend_coef$estimate * 12
```

The `trend()` coefficient is the estimated average change in mean monthly discharge (cfs) per month, after accounting for seasonality. Multiply by 12 for an annual rate.

```{r}
# The seasonal coefficients — all relative to the reference month
tidy(fit) %>% filter(stringr::str_detect(term, "season"))
```

Notice the output starts at `season()year2` — that's February. January (`year1`) is absent because it's the **reference category**, absorbed into the intercept. Every seasonal coefficient is the expected difference between that month and January, holding the trend constant. A positive value means that month runs higher than January on average; negative means lower.

## Model fit: observed vs. fitted

```{r}
fit %>%
  fabletools::augment() %>%
  ggplot(aes(x = ym)) +
  geom_line(aes(y = mean_flow, color = "Observed")) +
  geom_line(aes(y = .fitted,   color = "Fitted")) +
  scale_color_manual(values = c("Observed" = "gray50", "Fitted" = "steelblue")) +
  labs(
    title = "Observed vs. Fitted — TSLM",
    y     = "Mean Discharge (cfs)",
    x     = NULL,
    color = NULL
  )
```

Where does the model fit well? Where does it struggle? What does that tell you about the model's assumptions?

The residuals (observed − fitted) tell us exactly where the model is wrong. Large residuals = the model missed badly. Let's find the worst misses:

```{r}
fit %>%
  fabletools::augment() %>%
  as_tibble() %>%
  mutate(abs_resid = abs(.resid)) %>%
  arrange(desc(abs_resid)) %>%
  select(ym, mean_flow, .fitted, .resid) %>%
  head(10)
```

Large positive residuals (observed >> fitted) are typically flood months. Large negative residuals are drought months. The model only captures a linear trend and a fixed seasonal shape — it has no mechanism to anticipate extreme events, so those show up entirely in the residuals.

## Forecast

```{r}
fc <- fit %>%
  fabletools::forecast(h = "2 years")

fc %>%
  fabletools::autoplot(flow_monthly, level = c(80, 95)) +
  labs(
    title = "2-Year Forecast — Monthly Mean Streamflow",
    y     = "Mean Discharge (cfs)",
    x     = NULL
  )
```

The shaded bands are **prediction intervals** (80% and 95%).

**Questions**:

1. Does the forecast look reasonable? Does it match your intuition from the seasonal plot?
2. What does the *width* of the prediction intervals tell you?
3. What important things does this model NOT account for? What are its real-world limitations?

One sanity check for question 1 — compare the forecast values to the most recent observed values:

```{r}
# Last 12 months of observed data
flow_monthly %>% tail(12)

# First 12 months of the forecast
fc %>% as_tibble() %>% select(ym, .mean) %>% head(12)
```

The forecast should produce magnitudes in a similar ballpark to recent observations and preserve the seasonal shape.

For question 2: the prediction intervals widen as you forecast further into the future — uncertainty accumulates. A very wide interval isn't a model failure; it's an honest statement about how much we don't know. The 95% band is always wider than the 80% band by construction.

For question 3: the TSLM assumes the trend is linear and the seasonal shape is fixed and identical year to year. It cannot anticipate floods, droughts, dam operations, or any structural change in the watershed. It also can't tell you *which* spring will flood — only that spring tends to be high on average.

---

# Your turn

Choose one of the following tasks with the remaining time. Compare your results with classmates.

---

**Option 1 — Try a different gauge**

Find a USGS site number for a river you're interested in (use the [USGS Water Resources site](https://waterdata.usgs.gov/nwis) to find a site). Pull the data, convert to a monthly tsibble, and recreate at minimum: the seasonal plot, ACF, and decomposition.

Some Ohio gauges to try if you don't want to search:

| Site | Description |
|------|-------------|
| `04208504` | Cuyahoga River at Kent, OH |
| `04212100` | Grand River at Painesville, OH |
| `04208600` | Mahoning River at Youngstown, OH |

---

**Option 2 — Try a different model**

Go back to the monthly Cuyahoga data. Fit an ARIMA model instead of TSLM:

```{r eval=FALSE}
fit_arima <- flow_monthly %>%
  fabletools::model(
    arima = fable::ARIMA(mean_flow)
  )
```

Produce the same observed vs. fitted and forecast plots. How do the two models compare? Which do you prefer and why?

---

**Option 3 — Dig into the trend**

The decomposition suggested a possible trend. Is it statistically meaningful? Look at the `tidy(fit)` output again. What is the trend coefficient, its standard error, and its p-value? Write 2-3 sentences interpreting the result for a non-technical audience.

---

# Wrapping up

Today we covered:

- **`tsibble`** — the tidy time series data structure
- **Visualization** — time plots, seasonal plots, subseries plots, ACF
- **Decomposition** — separating trend, seasonality, and remainder with STL
- **Modeling** — fitting a simple trend + seasonality model and forecasting

The tidyverts ecosystem (`tsibble` + `feasts` + `fable`) has a lot more to offer. If you want to go deeper, Rob Hyndman's free textbook [Forecasting: Principles and Practice (3rd ed.)](https://otexts.com/fpp3/) is the best resource and uses exactly these packages.
