---
title: "Week 15: Introduction to Time Series Analysis"
output:
  html_document:
    df_print: paged
    toc: true
    toc_float: true
---

# Today's plan

1. Brief framing: what makes time series data special
2. Getting environmental data with `dataRetrieval`
3. The `tsibble`: a tidy time series object
4. Visualization (several kinds)
5. Decomposition: trend, seasonality, remainder
6. A simple model and forecast
7. Your turn

---

# Source and inspiration:

[Forecasting: Principles and Practice (3rd ed.)](https://otexts.com/fpp3/), by Rob J Hyndman and George Athanasopoulos.

---

# Packages

Today we're using the **tidyverts** ecosystem — a tidyverse-native set of packages for time series work.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(dataRetrieval)  # USGS water data
library(tsibble)        # tidy time series data structure
library(feasts)         # decomposition + ACF
library(ggtime)         # time series visualization (split from feasts 0.4.2)
library(fable)          # forecasting models
library(lubridate)
```

---

# What makes time series data special?

- Most statistical methods assume observations are **independent** of each other — knowing one data point tells you nothing about the next. 
- This assumption breaks down completely for data collected over time. Yesterday's river flow strongly predicts today's river flow. Last January's streamflow is a decent predictor of this January's. 
- This temporal dependence is called **autocorrelation**, and it can be a challenge AND a tool

## Why this matters

A LOT of data in environmental and spatial science is collected repeatedly over time: 
- streamflow measured daily for decades
- air quality readings every hour
- land surface temperature from satellite passes
- species counts from annual surveys 

The *temporal structure* in that data is **not** noise to be discarded. Intead, it can be useful information about the stability of the process we're studying. Time series methods let us extract this information.

Practical reasons for learning/using these methods: 
- If we ignore autocorrelation and apply standard regression or t-tests to time series data, the standard errors come out *too small*
- We end up thinking your results are more precise and statistically significant than reality
- Specialized time series methods exist BECAUSE they account for this structure rather than pretending it isn't there

*Sound familiar?*


## Stationarity

Many time series methods require the series to be **stationary** — meaning its statistical properties (mean, variance, correlation structure) don't change over time. A series with a long-term trend is *not* stationary: its mean is drifting. A series with strong seasonality is *not* stationary in the same way: the expected value in April is systematically higher than in August.

Is this a problem? Maybe, maybe not, but the solution is what makes time series analysis interesting: we can *decompose* the series into its non-stationary parts and analyze what remains.

## The decomposition framework

We will separate any time series into three components:

1. **Trend** — long-term increase or decrease (slow, persistent drift)
2. **Seasonality** — regular, repeating patterns tied to the calendar (annual cycles, weekly cycles, etc.)
3. **Remainder** — what's left; ideally, random noise with no remaining structure

Think of it like: *data = trend + seasonality + remainder*

Once decomposed, you can ask:
1. Is there evidence of a long-term trend? 
2. How large is the seasonal signal relative to that trend? 
3. Are the "leftovers" truly random, or is there still structure a model might capture?

### Quick discussion

Before we look at any data: if you were measuring streamflow on a river in Northeast Ohio, what kind of trend and seasonal pattern would you expect? Why?

> **With your partner (3 min):** Sketch the shape of a full year of streamflow on paper — just a rough curve, no numbers needed. Where are the peaks? The troughs? Then make a second sketch showing what you'd expect a 20-year trend to look like. Be ready to share one thing you disagreed on.

---

# Accessing today's data: USGS streamflow

The `dataRetrieval` package provides direct access to USGS water data. We'll look at streamflow on the **Cuyahoga River at Independence, OH**.

```{r}
site_no      <- "04208000"   # Cuyahoga River at Independence, OH
parameter_cd <- "00060"      # discharge, in cubic feet per second (cfs)

raw_flow <- dataRetrieval::readNWISdv(
  siteNumbers = site_no,
  parameterCd = parameter_cd,
  startDate   = "2000-01-01",
  endDate     = "2023-12-31"
)


head(raw_flow)
```

**Question**: What variables do you see?

The column name encodes USGS codes: `00060` = discharge parameter, `00003` = daily mean statistic. The `_cd` column is a data-quality flag. Let's look at what values it actually contains:

```{r}
raw_flow %>% count(X_00060_00003_cd)
```

`"A"` = approved (reviewed), `"P"` = provisional (subject to revision). For long-term analysis, approved data is preferable.

```{r}
# The package has a helper to rename columns to something readable
flow_clean <- raw_flow %>%
  dataRetrieval::renameNWISColumns() %>%
  dplyr::select(Date, Flow) %>%
  dplyr::filter(!is.na(Flow))

head(flow_clean)
```

How many observations do we have?

```{r}
nrow(flow_clean)
```

**Question**: Is that about what you'd expect for daily data from 2000–2023? Do the math.

```{r}
# expected days in our date range
expected_days <- as.numeric(as.Date("2023-12-31") - as.Date("2000-01-01")) + 1
expected_days

# how does that compare to what we actually have?
nrow(flow_clean)

# how many NAs were dropped during cleaning?
raw_flow %>% dataRetrieval::renameNWISColumns() %>% pull(Flow) %>% is.na() %>% sum()
```

Any shortfall between expected and actual is missing data — equipment outages, ice affecting the sensor, etc.

---

# The `tsibble`: a tidy time series object

A **tsibble** is like a tibble, but with an explicit time **index** and (optionally) a grouping **key**. The index tells the package how observations are ordered in time and infers the interval automatically.

```{r}
flow_ts <- flow_clean %>%
  tsibble::as_tsibble(index = Date)

flow_ts
```

The header line of the output above shows the interval tsibble detected — look for `[1D]`. Yes, that aligns with expectations: we pulled daily values. 

`tsibble` infers the interval automatically from the spacing of the index. If you had gaps, it would error out rather than silently skip them — which is useful.

---

# Visualization

## The basic time series plot

Here, we will use `feasts` and `ggtime` plotting functions rather than building everything from scratch in ggplot. As of feasts 0.4.2, visualization functions like `gg_season()` and `gg_subseries()` moved to the `ggtime` package, while analytical functions like `ACF()` and `STL()` stayed in `feasts`.
- Both packages understand the tsibble structure: the time index goes on the x-axis automatically, and lags and seasonal periods are handled correctly
- Every plot returns a regular ggplot object, so `+ labs()`, `+ theme()`, etc. all work normally

```{r}
flow_ts %>%
  feasts::autoplot(Flow) +
  labs(
    title = "Daily Streamflow — Cuyahoga River at Independence, OH",
    y     = "Discharge (cfs)",
    x     = NULL
  ) +
    theme_minimal() # see how you can add ggplot "stuff"?
```

**STOP**

What do you notice? Any obvious trend? Seasonality? Large spikes?

### Question - what do the large "spikes" represent?


There's a seasonal rhythm, but it's hard to characterize at daily resolution — the noise drowns it out.

 A long-term trend, if present, is (more or less) invisible here. Let's aggregate to monthly to make the signal easier to see.

## Aggregate to monthly

Daily data for 24 years is noisy. Monthly means will make seasonal patterns clearer and keep our models fast.

```{r}
flow_monthly <- flow_clean %>%
  dplyr::mutate(ym = tsibble::yearmonth(Date)) %>%
  dplyr::group_by(ym) %>%
  dplyr::summarise(mean_flow = mean(Flow, na.rm = TRUE)) %>%
  tsibble::as_tsibble(index = ym)

flow_monthly
```


Plot it - notice the call. What does `autoplot` do???

```{r}
flow_monthly %>%
  feasts::autoplot(mean_flow) +
  labs(
    title = "Monthly Mean Streamflow — Cuyahoga River at Independence, OH",
    y     = "Mean Discharge (cfs)",
    x     = NULL
  )
```

**How does this compare to the daily plot? What's easier or harder to see at monthly resolution?**

> **With your partner:** Modify the aggregation chunk to produce quarterly instead of monthly data, then re-run the plot. What do you gain and lose compared to monthly? Which aggregation would you use if your main question was about long-term trend vs. seasonal shape?
>
> *Hints:*
> - *`tsibble` has a `yearquarter()` function that works just like `yearmonth()` — swap it in the `mutate()` call*
> - *Rename `ym` to `yq` everywhere in the chunk so the variable name still makes sense*
> - *Update the `as_tsibble(index = ym)` line to match your new variable name*
> - *Then update the `autoplot()` call to use `yq` and `mean_flow`*

Try it using the below code block:

```{r}


```


The seasonal pattern is much clearer — regular annual peaks and troughs are no longer buried in noise. 

Trade-off: individual flood events are averaged away. For understanding long-term structure (seasonality, trend), monthly is appropriate. For flood frequency or magnitude, you want to stay with daily.

## Seasonal plot

This plots each year as its own line, all aligned on the same x-axis (month of year).

```{r}
flow_monthly %>%
  ggtime::gg_season(mean_flow, labels = "both") +
  labs(
    title = "Seasonal Plot — Monthly Mean Streamflow",
    y     = "Mean Discharge (cfs)"
  ) + theme_minimal()
```


Is this a useful plot???

> **With your partner:** Add a `filter(year(ym) >= 2010)` before the `gg_season()` call and re-run the plot. Does the seasonal shape look the same in the recent record as in the full record? Does any year stand out differently than it did before?


Try it here:

```{r}


```


**Questions**:

1. Which months consistently have the highest flow? Lowest? Can you tell?
2. How consistent is the seasonal pattern across years?
3. Can you identify any years that look unusual? What might explain them?

For question 1, let's compute this directly instead of just eyeballing:

```{r}
flow_monthly %>%
  as_tibble() %>%
  mutate(month = lubridate::month(ym, label = TRUE)) %>%
  group_by(month) %>%
  summarise(avg_flow = mean(mean_flow, na.rm = TRUE)) %>%
  arrange(desc(avg_flow))
```

For question 3, which individual years had the highest and lowest *annual* mean flow?

```{r}
flow_monthly %>%
  as_tibble() %>%
  mutate(year = lubridate::year(ym)) %>%
  group_by(year) %>%
  summarise(annual_mean = mean(mean_flow, na.rm = TRUE)) %>%
  arrange(desc(annual_mean))
```

Years at the top and bottom of this table are the ones that stand out as lines in the seasonal plot. The seasonal shape stays consistent(ish) year to year. The peak is almost always spring and the trough late summer — but the overall magnitude varies substantially.

## Subseries plot

Another way to look at seasonality: one panel per month, showing how that month's values change across years.

```{r}
flow_monthly %>%
  ggtime::gg_subseries(mean_flow) +
  labs(
    title = "Subseries Plot — Monthly Mean Streamflow",
    y     = "Mean Discharge (cfs)"
  )
```

The blue horizontal line in each panel is the mean for that month. What does this view tell you that the seasonal plot didn't?

The subseries plot lets you ask: within a specific month, is flow changing over time? The seasonal plot shows the annual shape well but makes month-specific trends invisible. Here's one way to make that question explicit:

```{r}
# Is March flow trending over time?
flow_monthly %>%
  as_tibble() %>%
  mutate(month = lubridate::month(ym),
         year  = lubridate::year(ym)) %>%
  filter(month == 3) %>%
  ggplot(aes(x = year, y = mean_flow)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "March Mean Streamflow Over Time",
       y = "Mean Discharge (cfs)", x = "Year")
```

**With your partner:** Each pair picks a different month — change `month == 3` to your chosen month and re-run the chunk. Does flow in your month show the same direction of change as March? Come back ready to say which month you picked and whether the trend surprised you.

## Autocorrelation

The ACF (Autocorrelation Function) measures how correlated each observation is with its own past values — at different lag distances.

What the ACF values actually mean, in plain English:

- The ACF value at **lag 1** answers: "If I know this month's flow, how well can I predict *next* month's flow?" A value near 1 means very well; near 0 means barely at all.
- The ACF value at **lag 12** answers: "How similar is this month's flow to the flow from *12 months ago* (same month, last year)?" A value near 1 means the same month repeats reliably year after year.
- An ACF value of **0** means knowing the past observation at that lag tells you nothing about the current value — the two are unrelated.
- ACF values are just **correlation coefficients** (ranging −1 to 1), so all the intuition you already have about correlation applies. The only difference is that both variables being correlated are the *same series*, just shifted in time.

```{r}
flow_monthly %>%
  feasts::ACF(mean_flow, lag_max = 36) %>%
  feasts::autoplot() +
  labs(title = "Autocorrelation Function — Monthly Mean Streamflow")
```

**Questions**:

1. What do the blue dashed lines represent?
2. At lag 1 the correlation is strongly positive — what does that mean in plain language?
3. What pattern do you see at lags 12, 24, 36? Why does that make sense given the data?

For question 1, the threshold is approximately ±2/√n. Let's compute it:

```{r}
n_months <- nrow(flow_monthly)
significance_threshold <- 2 / sqrt(n_months)
significance_threshold
```

Bars extending beyond this value are statistically distinguishable from zero — not just noise.

For question 3, we can extract the ACF values directly and look at the seasonal lags:

```{r}
flow_monthly %>%
  feasts::ACF(mean_flow, lag_max = 36) %>%
  as_tibble() %>%
  filter(as.numeric(lag) %in% c(1, 12, 24, 36))
```

The strong positive values at lags 12, 24, and 36 confirm annual seasonality: this month's flow is similar to the same month one, two, and three years ago. At lag 1, the positive correlation means high-flow months tend to follow high-flow months — the river has inertia, it doesn't jump randomly month to month.

---

# Decomposition

STL (Seasonal and Trend decomposition using Loess) separates our time series into its component parts. It's robust to outliers and flexible about the shape of the trend and seasonal components.

```{r}
dcmp <- flow_monthly %>%
  fabletools::model(
    stl = feasts::STL(mean_flow ~ trend(window = 21) + season(window = "periodic"))
  )

dcmp %>%
  fabletools::components() %>%
  feasts::autoplot() +
  labs(title = "STL Decomposition — Monthly Mean Streamflow")
```

Look at each panel:

- **data** — original series
- **trend** — long-term direction (smoothed out the seasonality)
- **season_year** — repeating annual cycle
- **remainder** — what's left over after removing trend and season

**With your partner (3 min):** Look at the four panels and answer these before we discuss as a class:

1. Is there a long-term trend in streamflow over this period? What direction?
2. How large is the seasonal component relative to the trend? (Hint: look at the y-axis scales)
3. The grey bars on the left side of each panel are deliberately the same height — what is their purpose?
4. What might explain large spikes in the remainder component?

For questions 1 and 2, we can extract the decomposition components and compare their ranges directly:

```{r}
dcmp_components <- dcmp %>% fabletools::components()

dcmp_components %>%
  as_tibble() %>%
  summarise(
    trend_range    = diff(range(trend,       na.rm = TRUE)),
    seasonal_range = diff(range(season_year, na.rm = TRUE)),
    remainder_range= diff(range(remainder,   na.rm = TRUE))
  )
```

Whichever range is largest dominates the variation in the series. In most river data, the seasonal range dwarfs the trend — knowing what month it is tells you more about flow than knowing what year it is.

For question 4, which dates have the largest remainder values (biggest departures from trend + season)?

```{r}
dcmp_components %>%
  as_tibble() %>%
  select(ym, remainder) %>%
  arrange(desc(abs(remainder))) %>%
  head(10)
```

You can look up those dates. They likely correspond to major flood events, drought years, or other unusual conditions. The grey bars in the decomposition plot serve as a visual scale reference: since all bars are the same physical height but span different y-axis ranges, you can visually compare how much variation each component actually accounts for.

---

# A simple model and forecast

We'll use a **Time Series Linear Model** (`TSLM`). Think of it as `lm()` but aware of the time structure. We'll include a linear trend term and monthly seasonal dummies.

```{r}
fit <- flow_monthly %>%
  fabletools::model(
    tslm = fable::TSLM(mean_flow ~ trend() + season())
  )

tidy(fit)
```

A few things to notice in the output before we interpret:

- `trend()` — a single numeric term counting months from the start of the series; its coefficient is the estimated change in flow per month
- `season()year2` through `season()year12` — one dummy variable per month, where the number is the month index (2 = February, 3 = March, … 12 = December)
- `season()year1` (January) is **missing on purpose** — it's the reference category, absorbed into the intercept; every other seasonal coefficient is the expected difference *relative to January*
- `p.value` tells you whether each term is statistically distinguishable from zero; most seasonal dummies will be highly significant because the months are genuinely different from each other

**Questions**:

1. What does the `trend()` coefficient tell you? What are its units?
2. There are seasonal coefficients — how would you interpret the one for `season()year2` (February)?
3. The model has a reference category for season — which month is it, and how can you tell?

```{r}
# Pull the trend coefficient and express it as monthly and annual change
trend_coef <- tidy(fit) %>% filter(term == "trend()")
trend_coef

# annual rate of change
trend_coef$estimate * 12
```

**Reading this output: **

- the `estimate` of **0.362** means the model estimates flow increases by about 0.36 cfs per month on average, after accounting for seasonality
- multiplied by 12, that's roughly **4.3 cfs per year**
- However, the `p.value` of **0.390** is well above any conventional threshold — so this trend is not statistically distinguishable from zero

**Implications and inference?**
The river may be trending upward slightly, but the data don't provide enough signal to be confident. The `std.error` of 0.420 is actually *larger* than the estimate itself, which reinforces that uncertainty.

```{r}
# The seasonal coefficients — all relative to the reference month
tidy(fit) %>% filter(stringr::str_detect(term, "season"))
```

Notice the output starts at `season()year2` — that's February. 

January (`year1`) is absent because it's the **reference category**, absorbed into the intercept. 

Every seasonal coefficient is the expected difference between that month and January, holding the trend constant. A positive value means that month runs higher than January on average; negative means lower.

## Model fit: observed vs. fitted

```{r}
fit %>%
  fabletools::augment() %>%
  ggplot(aes(x = ym)) +
  geom_line(aes(y = mean_flow, color = "Observed")) +
  geom_line(aes(y = .fitted,   color = "Fitted")) +
  scale_color_manual(values = c("Observed" = "gray50", "Fitted" = "red")) +
  labs(
    title = "Observed vs. Fitted — TSLM",
    y     = "Mean Discharge (cfs)",
    x     = NULL,
    color = NULL
  ) + theme_minimal()
```

Where does the model fit well? Where does it struggle? What does that tell you about the model's assumptions?

The residuals (observed − fitted) tell us exactly where the model is wrong. Large residuals = the model missed badly. Let's find the worst misses:

```{r}
fit %>%
  fabletools::augment() %>%
  as_tibble() %>%
  mutate(abs_resid = abs(.resid)) %>%
  arrange(desc(abs_resid)) %>%
  select(ym, mean_flow, .fitted, .resid) %>%
  head(10)
```

Large positive residuals (observed >> fitted) are typically flood months. Large negative residuals are drought months. The model only captures a linear trend and a fixed seasonal shape — it has no mechanism to anticipate extreme events, so those show up entirely in the residuals.

## Forecast

```{r}
fc <- fit %>%
  fabletools::forecast(h = "2 years")

fc %>%
  fabletools::autoplot(flow_monthly, level = c(80, 95)) +
  labs(
    title = "2-Year Forecast — Monthly Mean Streamflow",
    y     = "Mean Discharge (cfs)",
    x     = NULL
  )
```

The shaded bands are **prediction intervals** (80% and 95%).

**Questions**:

1. Does the forecast look reasonable? Does it match your intuition from the seasonal plot?
2. What does the *width* of the prediction intervals tell you?
3. What important things does this model NOT account for? What are its real-world limitations?

One sanity check for question 1 — compare the forecast values to the most recent observed values:

```{r}
# Last 12 months of observed data
flow_monthly %>% tail(12)

# First 12 months of the forecast
fc %>% as_tibble() %>% select(ym, .mean) %>% head(12)
```

The forecast should produce magnitudes in a similar ballpark to recent observations and preserve the seasonal shape.

For question 2: the prediction intervals widen as you forecast further into the future — uncertainty accumulates. A very wide interval isn't a model failure; it's an honest statement about how much we don't know. The 95% band is always wider than the 80% band by construction.

For question 3: the TSLM assumes the trend is linear and the seasonal shape is fixed and identical year to year. It cannot anticipate floods, droughts, dam operations, or any structural change in the watershed. It also can't tell you *which* spring will flood — only that spring tends to be high on average.

---

# A second model: ARIMA

TSLM is intuitive but makes strong assumptions — a perfectly linear trend, a seasonal pattern with fixed shape and size. ARIMA is more flexible. It doesn't assume a specific functional form; instead it learns the structure of the series from the data itself.

## What does ARIMA stand for?

ARIMA = **A**uto**R**egressive **I**ntegrated **M**oving **A**verage. Each word describes one ingredient:

- **AR — AutoRegressive**: The model predicts the next value using a weighted combination of *past values*. If last month's flow was high, the AR part says "expect this month's to be somewhat high too." The number of past values used is called *p*.

- **I — Integrated**: Many real-world series have trends, which violates a key assumption (called *stationarity*) that ARIMA needs. The "I" fixes this by working with *differences* between consecutive observations (this month minus last month) rather than the raw values. The number of times you difference is called *d*.

- **MA — Moving Average**: The model also adjusts predictions based on past *forecast errors* — how wrong was the model last time? If it over-predicted last month, it nudges down this month's forecast. The number of past errors used is called *q*.

The full notation is **ARIMA(p, d, q)**. For seasonal data, there's also a seasonal component written as **ARIMA(p, d, q)(P, D, Q)[m]**, where the uppercase letters apply to lags at the seasonal period *m* (12 for monthly data).

You do not need to pick p, d, q yourself. `fable::ARIMA()` selects them automatically by trying many combinations and choosing the one with the best information criterion (AICc).

## Fitting an ARIMA model

```{r}
fit_arima <- flow_monthly %>%
  fabletools::model(
    arima = fable::ARIMA(mean_flow)
  )
```

What order did it choose?

```{r}
report(fit_arima)
```

Read the model notation at the top of the output. For example, `ARIMA(1,1,1)(0,1,1)[12]` means:

- Non-seasonal part: 1 AR term, differenced once, 1 MA term
- Seasonal part: no seasonal AR, differenced once at lag 12, 1 seasonal MA term
- Seasonal period: 12 months

The exact order will vary — let the output tell you what it found.

## ARIMA forecast

```{r}
fc_arima <- fit_arima %>%
  fabletools::forecast(h = "2 years")

fc_arima %>%
  fabletools::autoplot(flow_monthly, level = c(80, 95)) +
  labs(
    title = "2-Year ARIMA Forecast — Monthly Mean Streamflow",
    y     = "Mean Discharge (cfs)",
    x     = NULL
  )
```

## Comparing TSLM and ARIMA side by side

We can fit both models in a single `model()` call and forecast them together:

```{r}
fit_both <- flow_monthly %>%
  fabletools::model(
    tslm  = fable::TSLM(mean_flow ~ trend() + season()),
    arima = fable::ARIMA(mean_flow)
  )

fit_both %>%
  fabletools::forecast(h = "2 years") %>%
  fabletools::autoplot(flow_monthly, level = NULL) +
  labs(
    title = "TSLM vs. ARIMA Forecast — Monthly Mean Streamflow",
    y     = "Mean Discharge (cfs)",
    x     = NULL
  )
```

`level = NULL` drops the shading so both forecasts are easier to compare as lines.

We can also compare how well each model fit the *historical* data using accuracy metrics:

```{r}
fit_both %>% fabletools::accuracy()
```

- **RMSE** (Root Mean Squared Error) and **MAE** (Mean Absolute Error) are both in the original units (cfs). Lower is better.
- **MAPE** (Mean Absolute Percentage Error) expresses error as a percentage of the observed value — useful for comparing across different scales.

Which model fits the historical data better? Does a better historical fit necessarily mean a better forecast?

---

# Your turn

Choose one of the following tasks with the remaining time. Compare your results with classmates.

---

**Option 1 — Try a different gauge**

Find a USGS site number for a river you're interested in (use the [USGS Water Resources site](https://waterdata.usgs.gov/nwis) to find a site). Pull the data, convert to a monthly tsibble, and recreate at minimum: the seasonal plot, ACF, and decomposition.

Some Ohio gauges to try if you don't want to search:

| Site | Description |
|------|-------------|
| `04208504` | Cuyahoga River at Kent, OH |
| `04212100` | Grand River at Painesville, OH |
| `04208600` | Mahoning River at Youngstown, OH |

---

**Option 2 — Dig deeper into the model comparison**

Look at the `accuracy()` output from `fit_both`. Which model has lower RMSE? Now look at the residuals of the ARIMA model — are they closer to random noise than the TSLM residuals?

```{r eval=FALSE}
# Residual plot for ARIMA
fit_arima %>%
  fabletools::augment() %>%
  ggplot(aes(x = ym, y = .resid)) +
  geom_line() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(title = "ARIMA Residuals", y = "Residual (cfs)", x = NULL)
```

A good model should have residuals that look like random noise — no obvious patterns, no trend, no seasonality left over. Does the ARIMA residual plot look more "random" than the TSLM residual plot? What does that tell you?

---

# Wrapping up

Today we covered:

- **`tsibble`** — the tidy time series data structure
- **Visualization** — time plots, seasonal plots, subseries plots, ACF
- **Decomposition** — separating trend, seasonality, and remainder with STL
- **Modeling** — TSLM (explicit trend + seasonal dummies) and ARIMA (learned from the data's autocorrelation structure), with forecasts and accuracy comparison

The tidyverts ecosystem (`tsibble` + `feasts` + `fable`) has a lot more to offer. If you want to go deeper, Rob Hyndman's free textbook [Forecasting: Principles and Practice (3rd ed.)](https://otexts.com/fpp3/) is the best resource and uses exactly these packages.
